---
title: "A brief introduction to tidy evaluation"
output: 
    github_document:
        html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = 'figures/')
```
```{r libs}
library(tidyverse)
theme_set(theme_classic())
```

## Tidy Eval: What & Why?

First, let's talk about variables. There are two kinds of variables we need to know about.

**`env-variables`** are variables that exist in our environment.
They're what you typically think of when you think of the word variable.
In R, you create them with the assignment operator (`<-`):
```{r}
name <- "Kelly"
numbers <- 1:3
ls()
```

Besides variables that you create, some packages include datasets that are also
available as `env-variables`. `iris` is a dataframe that's included in base R.
```{r}
head(iris)
```

**`data-variables`** are columns of dataframes. 
They aren't in our environment -- they live in their dataframes.

For example, `Species` is a `data-variable` in `iris`. 
The following code doesn't work:
```{r, error = TRUE}
print(Species)
```

In **base R**, we have to tell R which `env-variables` to access `data-variables` 
from using `$`:
```{r}
iris[iris$Species == "setosa" & iris$Sepal.Length > 5.5, ]
```

The **tidyverse** way allows us to avoid repeating the name of the dataframe:
```{r}
iris %>% filter(Species == "setosa", Sepal.Length > 5.5)
```

There's some magic going on under the hood of dplyr's `filter` function that 
lets us use `Species` and `Sepal.Length`, which are `data-variables`, 
as if they were `env-variables`.
Normally, R would complain that it can't find `Species`. 

**Tidy evaluation** is used in many tidyverse functions to
**delay evaluating the variables until later on inside the function code**.
This makes it slightly harder to use tidyverse functions inside our own functions,
but it's worth learning to write more reusable code!

### Example: dplyr functions

Let's say you're summarizing your data:
```{r}
iris %>%
    group_by(Species) %>%
    summarize(
        n = n(),
        min = min(Sepal.Length),
        max = max(Sepal.Length)
    )
```

And you realize you'd like to also summarize the `Sepal.Width` column.
You could copy-paste your code and change both instances of `Sepal.Length` to `Sepal.Width`:
```{r}
iris %>%
    group_by(Species) %>%
    summarize(
        n = n(),
        min = min(Sepal.Width),
        max = max(Sepal.Width)
    )
```

But now you have repetitive code. 
If you later decide you want to also calculate the mean of those columns,
you'll have to modify your code in two places -- 
or more if you kept on copy-pasting for more columns.

Instead, let's try writing a function that will summarize any column name we want.
```{r, error = TRUE}
var_summary <- function(data, var) {
    data  %>%
        group_by(Species) %>%
        summarize(n = n(),
                  min = min(var),
                  max = max(var))
}
iris %>%
    var_summary(Sepal.Length)
```

That didn't quite work because `Sepal.Length` is a `data-variable` that doesn't
exist in our environment, so R can't find it.  
We need to tell R to wait to evaluate the `var` argument until it gets to the 
`summarize` function.

We can do that by capturing the variable with the `enquo` function to create a 
`quosure` (`quoted closure`). 
A `quosure` captures a variable or expression so we can evaluate it later. 
Think of it like freezing in place. 
Using a special operator `!!` (pronounced `bang-bang`), we'll tell R to wait to 
evaluate our quosure until it's inside the `summarize` function.
```{r}
var_summary <- function(data, var) {
    var <- enquo(var)
    data  %>%
        group_by(Species) %>%
        summarize(n = n(),
                  min = min(!!var),
                  max = max(!!var))
}
iris %>%
    var_summary(Sepal.Length)
```

If you don't need to do anything with the `quosure` before unquoting it, you can use `!!enquo` all at once like this:
```{r}
var_summary <- function(data, var) {
    data  %>%
        group_by(Species) %>%
        summarize(n = n(),
                  min = min(!!enquo(var)),
                  max = max(!!enquo(var)))
}
iris %>%
    var_summary(Sepal.Length)
```

This pattern of using `!!enquo` is really common, 
so the tidyverse developers created a new operator called embracing `{{ }}` 
to do them both at the same time:
```{r}
var_summary <- function(data, var) {
  data %>% 
    group_by(Species) %>%
    summarize(n = n(), 
              min = min({{ var }}), 
              max = max({{ var }})
              )
}
```

Most of the time, embracing is what you'll want to use.

### Example: expressions

You can also pass more complicated expressions as arguments to functions.

```{r}
filter2 <- function(data, comparison1, comparison2) {
    data %>% 
        filter({{ comparison1 }}, {{ comparison2 }})
}
iris %>% filter2(Sepal.Length > Sepal.Width * 2.5, Species == "versicolor")
```

### Example: multiple arguments

You can pass any number of arguments to other functions with `...`
```{r}
summarize_vars <- function(.data, ...) {
  .data %>% 
    group_by(Species) %>%
    summarize_at(vars(...), mean)
}
iris %>% summarize_vars(Sepal.Length, Sepal.Width, Petal.Width)
```

Note:

- You don't need to use `!!enquo` or `{{ }}` when using `...` -- it just works!
- The triple-dots `...` has to be the last argument in the function signature.
- It's considered good practice to name your other arguments beginning with a `.` so users don't accidentally match them to other variables forwarded by `...` ([more details here](https://design.tidyverse.org/dots-prefix.html)).

## Exercises

### 1. Write a function to calculate the ratio of any two numeric variables

```{r}
# replace the X's to make this code work
calc_ratio <- function(data, var1, var2) {
    XXXXXX %>% 
        mutate(ratio = XXXXXXX / XXXXXXX)
}
# example:
#iris %>% calc_ratio(Sepal.Length, Sepal.Width)
#iris %>% calc_ratio(Petal.Width, Petal.Length)
```

### 2. Write a function to create a histogram that can use any column for `x`

Here's an example histogram where `x` is `Sepal.Length`:
```{r hist_example}
iris %>% 
    ggplot(aes(x = Sepal.Length, fill = Species)) +
    geom_histogram()
```

Create a function that takes a column to use for `x`:
```{r hist_exercise}
plot_hist <- function(data, x_col) {
    # write code here
}
plot_hist(iris, Sepal.Width)
plot_hist(iris, Petal.Length)
```

### 3. Write a function to create a boxplot with any aesthetics

```{r boxplots}
# write function here

# example:
#iris %>% plot_box(x = Species, y = Petal.Width, fill = Species)
#iris %>% plot_box(x = Species, y = Sepal.Length, color = Species)
```

### 4. Bonus: Write a function that names a new column based on the variable

How strong is your Google-foo?
```{r}
calc_mean <- function(data, var) {
    data %>% 
        summarize(XXXXXXXXX= mean({{ var }}))
}
# example:
#iris %>% calc_mean(Sepal.Length)
#>   mean_Sepal.Length
#> 1          5.843333
```


## Resources

- Vignette on tidy eval with dplyr (this document adapts parts of it): https://dplyr.tidyverse.org/dev/articles/programming.html
- Section on tidy eval in The Tidynomicon: http://tidynomicon.tech/nse.html#what-is-tidy-evaluation
- Tidy Eval book (in progress): https://tidyeval.tidyverse.org/
- Metaprogramming chapters of Advanced R: https://adv-r.hadley.nz/metaprogramming.html
